---
title: "Basics"
engine: knitr
---

<div style="height: 2em;"></div>


## Operators

### Assignment operators

```{r}
x <- 1
y <- 2
z <- x + y
```

`<-` assigns an object to a name. `->` is allowed too but is uncommon due to readability. `=` is also allowed but discouraged as it has more than one meaning.

Object names must start with a letter, and only contain:

- Letters (case-sensitive)
- Numbers
- Underscores
- Periods

Avoid the following single-character object names:

+-------------+----------------------------------------------+
| Character   | Special meaning(s)                           |
+=============+==============================================+
| `.`         | In formulae, it means all other variables.\  |
|             | Can also be used as a placeholder            |
+-------------+----------------------------------------------+
| c           | Combine                                      |
+-------------+----------------------------------------------+
| q           | Quit                                         |
+-------------+----------------------------------------------+
| t           | Matrix transpose                             |
+-------------+----------------------------------------------+
| C           | Set contrasts for a factor                   |
+-------------+----------------------------------------------+
| D           | Symbolic and algorithmic derivatives         |
+-------------+----------------------------------------------+
| F           | `FALSE`                                      |
+-------------+----------------------------------------------+
| I           | Treat an object as is                        |
+-------------+----------------------------------------------+
| T           | `TRUE`                                       |
+-------------+----------------------------------------------+

If you redefine these objects, it will mask their original meaning(s).


### Arithmetic operators

Common arithmetic operators include:

+------------+----------+
| Operator   | Meaning  |
+============+==========+
| `+`        | $+$      |
+------------+----------+
| `-`        | $-$      |
+------------+----------+
| `*`        | $\times$ |
+------------+----------+
| `/`        | $\div$   |
+------------+----------+
| `^`        | ^        |
+------------+----------+
| `sqrt()`   | √        |
+------------+----------+
| `log()`    | ln()     |
+------------+----------+


### Relational operators

Relational operators compare two values and return `TRUE`, `FALSE`, or `NA`. Common relational operators include:

+------------------+-------------------------------+-------------------------------+
| Operator         | Meaning                       | Example                       |
+==================+===============================+===============================+
| `>`, `>=`        | $>$, $\ge$                    | `1 > 2` is `FALSE`            |
+------------------+-------------------------------+-------------------------------+
| `<`, `<=`        | $<$, $\le$                    | `1 <= 2` is `TRUE`            |
+------------------+-------------------------------+-------------------------------+
| `==`, `!=`       | $=$, $\ne$                    | `1 == 2` is `FALSE`           |
+------------------+-------------------------------+-------------------------------+
| `%in%`           | Contained in a vector or list | `1 %in% c(1,2)` is `TRUE`     |
+------------------+-------------------------------+-------------------------------+


### Logical operators

Logical operators combine or negate logical ("Boolean") values. They also return `TRUE`, `FALSE`, or `NA`. Common logical operators include:

+--------------+--------------+-------------------------------+
| Operator     | Meaning      | Example                       |
+==============+==============+===============================+
| `&`          | AND          | `1>2 & 3<4` is `FALSE`        |
+--------------+--------------+-------------------------------+
| `|`          | OR           | `1>2 | 3<4` is `TRUE`         |
+--------------+--------------+-------------------------------+
| `!`          | NOT          | `!(1<2)` is `FALSE`           |
+--------------+--------------+-------------------------------+


## Conditional statements

Conditional statements (if a condition is met then do this, else do that) help us program our decisions according to logical values.

```{r}
cases <- 75

if (cases > 100) {
  print("Emergency")
} else if (cases > 50) {
  print("Warning")
} else {
  print("Caution")
}
```

## Loops

Conditional statements only evaluate one value at a time (`cases` above). Loops let us repeat the same decision rules across many values. For example, suppose we have the **vector** (defined in the [next chapter](06_objects.qmd)) as follows:

We can write a **for loop**:
```{r}
cases <- c(25, 75, 125)

for (x in cases) {
  if (x > 100) {
    print("Emergency")
  } else if (x > 50) {
    print("Warning")
  } else {
    print("Caution")
  }
}
```

For loops run for a predetermined number of times (e.g., once for each value in a vector). By contrast, **while loops** run until a condition is no longer met:
```{r}
cases <- 125

while (cases >= 0) {
  if (cases > 100) {
    print("Emergency")
  } else if (cases > 50) {
    print("Warning")
  } else {
    print("Caution")
  }
  cases <- cases - 25
}
```
