---
title: "Objects"
---

<div style="height: 2em;"></div>


## Objects vs. functions

[John M. Chambers](https://datascience.stanford.edu/people/john-chambers), the creator of the S programming language (the precursor to R) and a member of the R Core Team [@rfoundationcontributors], wrote in 2014 [@chambers2014objectoriented]:

> To understand computations in R, two slogans are helpful:
> 
> - Everything that exists is an object.
> - Everything that happens is a function call.

Numbers, strings, functions, and everything else (even expressions for evaluation) in R are **objects**. Objects are created, transformed, and combined by **function** calls, including operators (e.g., `x <- 1` or `1 + 2` or `1 < 2`).


## Data types

**Types** tell you how data is stored internally. Common data types include:

+-----------+----------------------------------------------------------------+
| Data type | Definition                                                     |
+===========+================================================================+
| Double    | Default numeric type (e.g., `1`, `1.5`)                        |
+-----------+----------------------------------------------------------------+
| Integer   | Whole number type (denoted by `L`, e.g., `1L`)                 |
+-----------+----------------------------------------------------------------+
| Logical   | Boolean (`TRUE` or `FALSE`)                                    |
+-----------+----------------------------------------------------------------+
| Character | Text (letters, numbers, symbols)*                              |
+-----------+----------------------------------------------------------------+
| List      | Collection of arbitrary objects, including of different types  |
+-----------+----------------------------------------------------------------+
\* *Strings (e.g., `"apple"`) are sequences of characters enclosed in quotes (`''` or `""`).*

You can find an object’s type using `typeof()`. To check whether an object is a particular data type, use `is.integer()`, `is.character()`, and so on. To convert objects to other data types (if it is possible, such as 5 ↔ “5” is okay but 5 ↔ “apple” is not), use `as.integer()`, `as.character()`, and so on.


## Classes

**Classes** tell you how data behaves; objects belonging to a class share properties. There are only a few data types but more classes, including:

- Numeric
- Integer
- Logical
- Character
- Array
- Matrix
- List
- Data frame
- Factor
- Date
- Function

You can find an object's class using `class()`.

For example, `1.2` is stored as type `double` and behaves as class `numeric`. And `as.Date("2026-07-15")` is also stored as type `double` but behaves as class `Date`.


## Data structures

**Data structures** describe how data is organized. They can vary by data type and dimension.

![Data structures in R [@grolemundhandson]](assets/images/r_datastructures.png){width=400px}

### Vectors

```{r}
# Construct vectors using c()
my_vector <- c(1, 2, 3)
my_vector <- c("1", "2", "3")
my_vector

# Find elements by indexing (note: unlike Python, negative indices
# remove elements, rather than look backwards)
my_vector[2]

# You can name each element
my_vector = c(
  cases_2020 = 25,
  cases_2021 = 125
)
my_vector

# Find elements by indexing their name
my_vector["cases_2020"]

# If you try to construct vectors using elements of different types,
# R will try to coerce the elements to match. In the
# following example, TRUE becomes 1 (FALSE would be 0).
c(TRUE, 1)

# A sequence is a vector whose elements follow a regular progression
0:6
seq(0, 6, by = 2)

# A vector with 1 element can just be passed as is
1 == c(1)
```

### Matrices

```{r}
# A matrix is constructed vertically, from left to right
matrix(c(1,2,3,4,5,6,7,8), nrow = 4, ncol = 2)
```

### Arrays

```{r}
# An array is constructed vertically, from left to right, and top to
# bottom; it is not easy to visualize arrays!
array(c(1:4, 11:14, 21:24), dim = c(2,2,3))
```

### Lists

```{r}
# Construct lists using list(); lists can contain elements of the
# same or different data types
my_list <- list(1, 2, 3)
my_list <- list(TRUE, 1, "1")
my_list <- list("A", c(1, 2))
my_list

# Find elements and sub-elements by indexing
my_list[1]
my_list[[1]]
my_list[2]
my_list[[2]]
my_list[[2]][2]

# You can name each element
my_list = list(
  country = "Canada",
  cases_2020 = 25
)
my_list

# Find elements and sub-elements by indexing their name
my_list$cases_2020
```

### Data frames

Data frames are lists of vectors that have the same length. They are the most useful data structure for analyses because they are tabular -- just like spreadsheets.

```{r}
# Construct data frames using data.frame()
my_df <- data.frame(
  year = c(2020, 2021),
  cases = c(25, 125)
)
my_df

# Find columns and their elements by indexing their row, column
my_df[,2]
my_df[1,2]

# Find columns and their elements by indexing their name
my_df$cases
my_df$cases[1]

# Display the structure of an object
str(my_df)

# Display the summary of an object
summary(my_df)
```

::: {.callout-tip}
## **Tip**
A data frame has type `list` (because it is a list of equal-length vectors), but it has class `data.frame`.

```{r}
df <- data.frame(x = 1:3)
typeof(df)
class(df)
```
:::


## Functions

A function is an object that stores instructions for reuse.

```{r}
# Construct functions using function() {}
my_function <- function(a, b) {
  return(a + b)
}
my_function(a=1, b=2)  # Or just my_function(1, 2)
```

A function has **parameters** (`a` and `b`, above). When calling a function, you pass values for the parameters called **arguments** (`1` and `2`, above). This is where the `=` symbol is appropriate (why we avoid using `=` as an assignment operator).

Functions return values, though for some functions like plotting and writing, the returned value is often invisible. Functions do not need to have an explicit `return` statement like in the example above. R returns the value of the last evaluated expression anyway, so a `return` statement is unnecessary, but it is good to be explicit.
```{r}
# Equivalent function
my_function <- function(a, b) {
  a + b
}
```


::: {.content-visible when-format="html"}
## References
:::
